---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
This assumes the activity.csv file is in the current working directory. This will convert the date column to date type.
```{r}
data<- read.csv("activity.csv", header=T, sep=',', colClasses=c("numeric", "character", "numeric")) 
library(ggplot2)
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- as.factor(data$interval)
head(data)
str(data)
```


## What is mean total number of steps taken per day?
First, I will aggregate the data to get the sum of steps per date. I am removing NA values.  
```{r}
data1 <- aggregate(steps ~ date, data, sum, na.rm = TRUE)
head(data1)
```
I will then use the ggplot function to create a histogram.
```{r}
ggplot(data1, aes(x = steps)) + 
  geom_histogram(fill = "blue", binwidth = 500) + 
  labs(title="Steps Taken per Day", 
       x = "Total Number of Steps per Day", y = "Frequency (number of times completed)")
```

Finally, I will computer the mean and median of the total number of steps.
```{r}
mean_data1 <- mean(data1$steps)
med_data1 <- median(data1$steps)
mean_data1
med_data1
```
## What is the average daily activity pattern?
First, I will aggregate the data using the mean of steps over the intervals. 
```{r}
data2 <- aggregate(steps ~ interval, data, mean, na.rm=TRUE)
head(data2)
```
I will then use the plot function to plot the average number of steps over the intervals.
```{r}
plot(data2$steps, type = "l", main = "Average Number of steps taken per time interval", ylab = "Average number of steps", xlab = "Time Interval")
```

Here I will determine the the interval with the max number of steps:
```{r}
data2[which.max(data2$steps),]
```
## Imputing missing values
First, I will find the number of NA values
```{r}
total <- sum(is.na(data))
total
```
Then I will aggregate the data using the mean function. I will rename the columns of the new dataset (mean_data)
```{r}
mean_data <- aggregate(steps ~ interval, data, mean)
colnames(mean_data) <- c("interval", "mean_steps")
```
I will now merge the data with the new dataset created above. 
```{r}
data3 <- merge(data, mean_data)
```
Next, I will create a new column and if steps is NA, fill in the new column with the corresponding mean step value. Otherwise, fill it in with the current step value.
```{r}
data3$updated_steps <- ifelse(is.na(data3$steps), 
                                  data3$mean_steps, data3$steps)
```
I will aggregate the new dataset using the updated_Steps column and the sum function
```{r}
data4 <- aggregate(updated_steps ~ date, data3, sum)
head(data4)
```
Here, I will create the histogram
```{r}
ggplot(data4, aes(x = updated_steps)) + 
  geom_histogram(fill = "green", binwidth = 500) + 
  labs(title="Steps Taken per Day", 
       x = "Total Number of Steps per Day (NAs filled in)", y = "Frequency (number of times completed)")
```

Finally, I will compute the mean and median of the new dataset
```{r}
mean_data4 <- mean(data4$updated_steps)
med_data4 <- median(data4$updated_steps)
mean_data4
med_data4
```
These numbers are very close to the original dataset but they are slightly different. The median is the value that has increased with the new dataset.
## Are there differences in activity patterns between weekdays and weekends?
First I will create a character vector from the known dates in the data set
```{r}
weekday <- tolower(weekdays(data3$date))
```
Create a new column that displays the type of day using if else function
```{r}
data3$type_of_day <- as.factor(ifelse(weekday == "saturday" | weekday == "sunday", 
                                     "weekend", "weekday"))
```
Aggregate the data using the mean. Be sure to keep the type of day column!
```{r}
data5 <- aggregate(steps ~ interval + type_of_day, data3, mean)
```
Load the lattice package. Use xyplot to generate two graphs, one for weekends and one for weekdays.
```{r}
library(lattice)
xyplot(steps ~ interval | type_of_day, data5, type = "l", layout = c(1, 2), 
       xlab = "Interval", ylab = "Number of steps" )
```